Finch: Sparse and Structured Array Programming with Control Flow

### 2024-04-06 sections/tensors.tex
fix

\section{Bridging Looplets and Finch: The Tensor Interface}
%
%The Finch language provides descriptions of computations that iterate over a subset of a regular grid that is lexicographically ordered.
%
%Based on our description of Looplets, the reader might believe that compilation of a Finch program simply involves simply replacing for loops over a range with for loops over iterators.
%
%However, Finch's data structures and control flow are sufficiently complex that several challenges exist.
%
Arrays use multiple dimensions to organize data with respect to orthogonal concepts.
%
Thus, the Finch language supports multi-dimensional arrays.
%
Unfortunately, the Looplet abstraction is best suited towards iterators over a single dimension.
%
Our level abstraction provides a bridge between the single dimensional iterators created from looplets and the mutli-dimensional fiber-tree abstractions common to tensor compilers.
% 
This bridge must address three challenges.
%
First, while looplets represent an instance of an iterator over a tensor, we may access the same tensor twice with different indices.
%
Thus, the $unfurl$ function creates separate looplet nests for each iterator.
%
Next, since Finch programs go beyond just single einsums, they may read and write to the same data at different times.
%
The $declare$, $freeze$, and $thaw$ functions provide machinery to manage transition between these states.
%
Finally, we must be able to write looplet nests that modify tensors, as well as reading them.
%
The $assemble$ function manages the allocation of new data in the tensor.

@@

% \willow{After making changes to the introduction of this section, some of the following text feels redundant}
% To build our bridge, we embrace a set of abstractions: level formats/Fiber Trees, iteration context dependent instantiation of iterators, and tensor life cycles.
% %
% Our first abstraction mostly already exists in the literature: a manner of specifying a data structure for a multi-dimensional tensor out of data structures for single dimensional tensors
% %
% We recapitulate the essential details here.
% %
% Our next two abstractions add to to the first by providing a mechanism to use data structures generated by the first abstraction in a greater variety of contexts while maintaining per-dimension encapsulation of array data structures.
% %
% We introduce an interface to instatiate iterators in a variety of contexts in our programs and we introduce the lifecycle interface to manage when we read and write to multi-dimensional iterators.
% %
% These interfaces add to the level abstraction, expanding the types of data that they can express via mapping to looplets and expanding the contexts in which they can be used.
% %
% Previous efforts to compile a greater variety of sparse array programs left these bridges untouched ~\cite{henry_compilation_2021, won2023unified, senanayake2020sparse}.

@@

Our simplified view of a level is enabled by our use of looplets to represent
the structure within each fiber.
%
In fact, our level interface requires only 5
highly general operations, described below.

@@

\begin{wrapfigure}{r}{.4\textwidth}
    \vspace{-12pt}
    \centering
    \scriptsize
    \begin{tabular}{|c|c|c|c|l|}
        \hline
        \rothead{Sparse} & \rothead{Blocked} & \rothead{Runs} & \rothead{Singular } & \begin{tabular}{@{}l@{}}\textbf{Corresponding} \\ \textbf{Format}\end{tabular} \\ \hline
        &  &  &  & Dense \\ \hline
        &  &  & $\checkmark$ & n/a \\ \hline
        &  & $\checkmark$ &  & DenseRLE \\ \hline
        &  & $\checkmark$ & $\checkmark$ & n/a \\ \hline
        & $\checkmark$ &  &  & n/a \\ \hline
        & $\checkmark$ &  & $\checkmark$ & n/a \\ \hline
        & $\checkmark$ & $\checkmark$ &  & n/a \\ \hline
        & $\checkmark$ & $\checkmark$ & $\checkmark$ & n/a \\ \hline
       $\checkmark$ &  &  &  & Sparse \\ \hline
       $\checkmark$ &  &  & $\checkmark$ & SparsePinpoint \\ \hline
       $\checkmark$ &  & $\checkmark$ &  & SparseRLE \\ \hline
       $\checkmark$ &  & $\checkmark$ & $\checkmark$ & SparseInterval \\ \hline
       $\checkmark$ & $\checkmark$ &  &  & SparseVBL \\ \hline
       $\checkmark$ & $\checkmark$ &  & $\checkmark$ & SparseBand \\ \hline
       $\checkmark$ & $\checkmark$ & $\checkmark$ &  & n/a \\ \hline
       $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & n/a \\ \hline
    \end{tabular}
    \caption{All combinations of our 4 structural properties and the
    corresponding formats we have chosen to represent them. Not all combinations
    are relevant. Note that blocks and runs need not be considered together
    because we must store a run length for each run, so there isn't a
    significant storage benefit to combining them. Blocks and singletons only
    make sense in the contex of sparsity.}
    \vspace{-24pt}
    %We omit such combinations from the exhaustive table.
    \label{tab:TypesOfStructure}
\end{wrapfigure}

@@

@@

The leaf level stores the actual entries of the tensor. In most cases, it is
sufficient to store each entry at a separate position in a vector. This is accomplished by the \textbf{ElementLevel}. However, when all of the
values are the same, an additional optimization can be made by storing the
identical value only once. In this work, we introduce the concept of a \textbf{pattern}
level to handle this binary case. Pattern has a fill value of $false$, and returning $true$ for all ``stored'' values. The pattern level allows us to easily represent unweighted graphs or other boolean matrices.

    \hline
    \textbf{Sequentially Constructed Levels} \\
    \hline
    \textbf{Dense}:
    The dense format is the simplest format, mapping $fiber(l, p)[i] \rightarrow fiber(l.lvl, p \times l.shape + i)$.
    %
    This format is used to store dense data and is often a convenient format for the root level of a tensor.
    %
    Due to its simplicity, freezing and thawing the level are no-ops. \\
    \textbf{DenseRLE}:
    Used to represent runs of repeated values, storing two vectors, $right$ and $ptr$, with $qth$ run in the $p^th$ subfiber starting and ending at $right[ptr[p] + q]$ and $right[ptr[p] + q + 1] - 1$, respectively.
    %
    A challenge arises for this level: it is difficult to merge duplicate runs.
    %
    Such a scenario might arise when merging runs of subfibers of length 3, representing colors in an image.
    %
    Ideally, we would be able to detect duplicate subfibers and merge them on the fly, but we cannot determine which subfibers are equal because we cannot read them while the sublevel is in update-only mode.
    %
    Instead, we merge the duplicates during the freeze phase.
    %
    We $freeze$ the sublevel, $declare$ a separate sublevel $buf$ as a buffer to store the deduplicated subfibers, and then compare each of the subfibers in the main level, copying the deduplicated subfibers into the buffer. \\
    %
    %It also stores a duplicate sublevel, $buf$, for deduplication. \\
    \textbf{SparseList}:
    The simplest sparse format, used to construct popular formats like CSR, CSC, DCSR, DCSC, and CSF.
    %
    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$. \\
    \textbf{SparsePinpoint}:
    Similar to SparseList, but only one nonzero in each subfiber, eliminating the need for the $ptr$ field.
    %
    It stores a vector $idx$, such that $idx[p]$ is the nonzero index in the subfiber at position $p$. \\
    \textbf{SparseRLE}:
    Similar to DenseRLE level, but because runs are sparse, we must also store the start of each run.
    %
    It stores three vectors $left$, $right$, and $ptr$, such that the $qth$ run in the $p^th$ subfiber begins and ends at $left[ptr[p] + q]$ and $right[ptr[p] + q]$, respectively.
    %
    Like DenseRLE, it also stores a duplicate sublevel, $buf$, for deduplication. \\
    \textbf{SparseInterval}:
    Similar to SparseRLE, but only stores one run per subfiber, eliminating the need for the $ptr$ field.
    %
    This level does not deduplicate as it cannot store intermediate results with more than one run.
    %
    It stores two vectors, such that the run in subfiber $p$ begins and ends at $left[p]$ and $right[p]$ respectively. \\
    \textbf{SparseVBL}: Used to represent blocked data. It stores three vectors, $idx$, $ptr$, and $ofs$, such that $ofs[ptr[p] + q]:ofs[ptr[p] + q + 1] - 1$ are the subpositions of block $q$ ending at index $idx[ptr[p] + q]$ in the subfiber at position $p$. \\
    \textbf{SparseBand}: Similar to SparseVBL, but stores only one block per subfiber, eliminating the need for the $ptr$ field. It stores two vectors $idx$ and $ofs$, such that $ofs[p]:ofs[p + 1] - 1$ are the subpositions of the block ending at $idx[p]$ in subfiber $p$. \\

    \hline
    \textbf{Nonsequentially Constructed Levels} \\
    \hline
    \textbf{SparseHash}:
    The sparse hash format uses a hash table to store the locations of nonzeros, and sorts the unique indices for iteration during the freeze phase.
    %
    This allows for efficient random access, but not incremental construction, as the freeze phase runs in time proportional to the number of nonzeros in the entire level.
    %
    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$. 
    %
    Also stores a hash table $tbl$ for construction and random access in the level. \\
    \textbf{SparseBytemap}
    The SparseBytemap format uses a bytemap to store which locations have been written to.
    %
    Unlike the SparseHash format, the bytemap assembles the entire space of possible subfibers.
    %
    This accelerates random access in the format, but requires a high memory overhead.
    %
    Because we don't want to reallocate all of the memory in each iteration, the declaration of this format instead re-assembles only the dirty locations in the tensor.
    %
    This format is analogous to the default workspace format used by TACO.
    %
    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$.
    %
    These vectors are used to collect dirty locations.
    %
    It also stores $tbl$, a dense array of booleans such that $tbl[shape * p + i]$ is true when there is a nonzero at index $i$ in the subfiber at position $p$. \\
    \hline
    \textbf{Leaf Levels} \\
    \hline
    \textbf{Element}:
    The element level uses an array $val$ to store a value for each position $p$. The zero (fill) value is configurable.\\
    \textbf{Pattern}:
    The pattern level statically represents a leaf level with a fill value of $false$ and whose stored values are all $true$. \\
    \hline
    \textbf{Scalars} \\
    \hline
    \textbf{Scalar}:
    A dense scalar that, unlike a variable, supports reduction.  \\
    \textbf{SparseScalar}:
    A scalar with a dirty bit which specializes on the fill value when it occurs.\\
    \textbf{EarlyBreakScalar}:
    A scalar which triggers early breaks in reductions whenever an annihilator is encountered.\\
    \hline
    \end{tabular}
    \vspace{8pt}
    \caption{The main level formats supported by Finch. Note that all non-leaf
    levels store a the dimension of the subfibers and a child level. Since we
    must be able to handle the case where a sublevel is not stored because a
    parent level is sparse, all of Finch's sparse formats use a dirty bit during
    writing to determine whether the sublevel has been modified from it's
    default fill value and thus, whether it needs to be stored.}
    \label{table:formats}
\end{table}

>>>>

\section{Bridging Looplets and Finch: The Tensor Interface}
%
%The Finch language provides descriptions of computations that iterate over a subset of a regular grid that is lexicographically ordered.
%
%Based on our description of Looplets, the reader might believe that compilation of a Finch program simply involves simply replacing for loops over a range with for loops over iterators.
%
%However, Finch's data structures and control flow are sufficiently complex that several challenges exist.
%
Arrays use multiple dimensions to organize data with respect to orthogonal concepts.
%
Thus, the Finch language supports multi-dimensional arrays.
%
Unfortunately, the Looplet abstraction is best suited towards iterators over a single dimension.
%
Our level abstraction provides a bridge between the single dimensional iterators created from Looplets and the multi-dimensional fiber-tree abstractions common to tensor compilers.
% 
This bridge must address three challenges.
%
First, while Looplets represent an instance of an iterator over a tensor, we may access the same tensor twice with different indices.
%
Thus, the $unfurl$ function creates separate looplet nests for each iterator.
%
Next, since Finch programs go beyond just single Einsums, they may read and write to the same data at different times.
%
The $declare$, $freeze$, and $thaw$ functions provide machinery to manage transition between these states.
%
Finally, we must be able to write looplet nests that modify tensors, as well as reading them.
%
The $assemble$ function manages the allocation of new data in the tensor.

@@

% \willow{After making changes to the introduction of this section, some of the following text feels redundant}
% To build our bridge, we embrace a set of abstractions: level formats/Fiber Trees, iteration context dependent instantiation of iterators, and tensor life cycles.
% %
% Our first abstraction mostly already exists in the literature: a manner of specifying a data structure for a multi-dimensional tensor out of data structures for single dimensional tensors
% %
% We recapitulate the essential details here.
% %
% Our next two abstractions add to to the first by providing a mechanism to use data structures generated by the first abstraction in a greater variety of contexts while maintaining per-dimension encapsulation of array data structures.
% %
% We introduce an interface to instatiate iterators in a variety of contexts in our programs and we introduce the lifecycle interface to manage when we read and write to multi-dimensional iterators.
% %
% These interfaces add to the level abstraction, expanding the types of data that they can express via mapping to Looplets and expanding the contexts in which they can be used.
% %
% Previous efforts to compile a greater variety of sparse array programs left these bridges untouched ~\cite{henry_compilation_2021, won2023unified, senanayake2020sparse}.

@@

Our simplified view of a level is enabled by our use of Looplets to represent
the structure within each fiber.
%
In fact, our level interface requires only 5
highly general operations, described below.

@@

\begin{wrapfigure}{r}{.4\textwidth}
    \vspace{-12pt}
    \centering
    \scriptsize
    \begin{tabular}{|c|c|c|c|l|}
        \hline
        \rothead{Sparse} & \rothead{Blocked} & \rothead{Runs} & \rothead{Singular } & \begin{tabular}{@{}l@{}}\textbf{Corresponding} \\ \textbf{Format}\end{tabular} \\ \hline
        &  &  &  & Dense \\ \hline
        &  &  & $\checkmark$ & n/a \\ \hline
        &  & $\checkmark$ &  & DenseRLE \\ \hline
        &  & $\checkmark$ & $\checkmark$ & n/a \\ \hline
        & $\checkmark$ &  &  & n/a \\ \hline
        & $\checkmark$ &  & $\checkmark$ & n/a \\ \hline
        & $\checkmark$ & $\checkmark$ &  & n/a \\ \hline
        & $\checkmark$ & $\checkmark$ & $\checkmark$ & n/a \\ \hline
       $\checkmark$ &  &  &  & Sparse \\ \hline
       $\checkmark$ &  &  & $\checkmark$ & SparsePinpoint \\ \hline
       $\checkmark$ &  & $\checkmark$ &  & SparseRLE \\ \hline
       $\checkmark$ &  & $\checkmark$ & $\checkmark$ & SparseInterval \\ \hline
       $\checkmark$ & $\checkmark$ &  &  & SparseVBL \\ \hline
       $\checkmark$ & $\checkmark$ &  & $\checkmark$ & SparseBand \\ \hline
       $\checkmark$ & $\checkmark$ & $\checkmark$ &  & n/a \\ \hline
       $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & n/a \\ \hline
    \end{tabular}
    \caption{All combinations of our 4 structural properties and the
    corresponding formats we have chosen to represent them. Not all combinations
    are relevant. Note that blocks and runs need not be considered together
    because we must store a run length for each run, so there isn't a
    significant storage benefit to combining them. Blocks and singletons only
    make sense in the context of sparsity.}
    \vspace{-24pt}
    %We omit such combinations from the exhaustive table.
    \label{tab:TypesOfStructure}
\end{wrapfigure}

@@

    \hline
    \textbf{Sequentially Constructed Levels} \\
    \hline
    \textbf{Dense}:
    The dense format is the simplest format, mapping $fiber(l, p)[i] \rightarrow fiber(l.lvl, p \times l.shape + i)$.
    %
    This format is used to store dense data and is often a convenient format for the root level of a tensor.
    %
    Due to its simplicity, freezing and thawing the level are no-ops. \\
    \textbf{DenseRLE}:
    Used to represent runs of repeated values, storing two vectors, $right$ and $ptr$, with $q^{th}$ run in the $p^{th}$ subfiber starting and ending at $right[ptr[p] + q]$ and $right[ptr[p] + q + 1] - 1$, respectively.
    %
    A challenge arises for this level: it is difficult to merge duplicate runs.
    %
    Such a scenario might arise when merging runs of subfibers of length 3, representing colors in an image.
    %
    Ideally, we would be able to detect duplicate subfibers and merge them on the fly, but we cannot determine which subfibers are equal because we cannot read them while the sublevel is in update-only mode.
    %
    Instead, we merge the duplicates during the freeze phase.
    %
    We $freeze$ the sublevel, $declare$ a separate sublevel $buf$ as a buffer to store the deduplicated subfibers, and then compare each of the subfibers in the main level, copying the deduplicated subfibers into the buffer. \\
    %
    %It also stores a duplicate sublevel, $buf$, for deduplication. \\
    \textbf{SparseList}:
    The simplest sparse format, used to construct popular formats like CSR, CSC, DCSR, DCSC, and CSF.
    %
    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $q^{th}$ nonzero in the subfiber at position $p$. \\
    \textbf{SparsePinpoint}:
    Similar to SparseList, but only one nonzero in each subfiber, eliminating the need for the $ptr$ field.
    %
    It stores a vector $idx$, such that $idx[p]$ is the nonzero index in the subfiber at position $p$. \\
    \textbf{SparseRLE}:
    Similar to DenseRLE level, but because runs are sparse, we must also store the start of each run.
    %
    It stores three vectors $left$, $right$, and $ptr$, such that the $q^{th}$ run in the $p^{th}$ subfiber begins and ends at $left[ptr[p] + q]$ and $right[ptr[p] + q]$, respectively.
    %
    Like DenseRLE, it also stores a duplicate sublevel, $buf$, for deduplication. \\
    \textbf{SparseInterval}:
    Similar to SparseRLE, but only stores one run per subfiber, eliminating the need for the $ptr$ field.
    %
    This level does not deduplicate as it cannot store intermediate results with more than one run.
    %
    It stores two vectors, such that the run in subfiber $p$ begins and ends at $left[p]$ and $right[p]$ respectively. \\
    \textbf{SparseVBL}: Used to represent blocked data. It stores three vectors, $idx$, $ptr$, and $ofs$, such that $ofs[ptr[p] + q]:ofs[ptr[p] + q + 1] - 1$ are the subpositions of block $q$ ending at index $idx[ptr[p] + q]$ in the subfiber at position $p$. \\
    \textbf{SparseBand}: Similar to SparseVBL, but stores only one block per subfiber, eliminating the need for the $ptr$ field. It stores two vectors $idx$ and $ofs$, such that $ofs[p]:ofs[p + 1] - 1$ are the subpositions of the block ending at $idx[p]$ in subfiber $p$. \\

    \hline
    \textbf{Nonsequentially Constructed Levels} \\
    \hline
    \textbf{SparseHash}:
    The sparse hash format uses a hash table to store the locations of nonzeros, and sorts the unique indices for iteration during the freeze phase.
    %
    This allows for efficient random access, but not incremental construction, as the freeze phase runs in time proportional to the number of nonzeros in the entire level.
    %
    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $q^{th}$ nonzero in the subfiber at position $p$. 
    %
    Also stores a hash table $tbl$ for construction and random access in the level. \\
    \textbf{SparseBytemap}
    The SparseBytemap format uses a bytemap to store which locations have been written to.
    %
    Unlike the SparseHash format, the bytemap assembles the entire space of possible subfibers.
    %
    This accelerates random access in the format, but requires a high memory overhead.
    %
    Because we don't want to reallocate all of the memory in each iteration, the declaration of this format instead re-assembles only the dirty locations in the tensor.
    %
    This format is analogous to the default workspace format used by TACO.
    %
    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $q^{th}$ nonzero in the subfiber at position $p$.
    %
    These vectors are used to collect dirty locations.
    %
    It also stores $tbl$, a dense array of booleans such that $tbl[shape * p + i]$ is true when there is a nonzero at index $i$ in the subfiber at position $p$. \\
    \hline
    \textbf{Leaf Levels} \\
    \hline
    \textbf{Element}:
    The element level uses an array $val$ to store a value for each position $p$. The zero (fill) value is configurable.\\
    \textbf{Pattern}:
    The pattern level statically represents a leaf level with a fill value of $false$ and whose stored values are all $true$. \\
    \hline
    \textbf{Scalars} \\
    \hline
    \textbf{Scalar}:
    A dense scalar that, unlike a variable, supports reduction.  \\
    \textbf{SparseScalar}:
    A scalar with a dirty bit which specializes on the fill value when it occurs.\\
    \textbf{EarlyBreakScalar}:
    A scalar which triggers early breaks in reductions whenever an annihilator is encountered.\\
    \hline
    \end{tabular}
    \vspace{8pt}
    \caption{The main level formats supported by Finch. Note that all non-leaf
    levels store a the dimension of the subfibers and a child level. Since we
    must be able to handle the case where a sublevel is not stored because a
    parent level is sparse, all of Finch's sparse formats use a dirty bit during
    writing to determine whether the sublevel has been modified from it's
    default fill value and thus, whether it needs to be stored.}
    \label{table:formats}
\end{table}

\subsubsection{Scalars}
Because leaf levels are geared towards representing multiple leaves, we also introduce a much simpler Scalar format to represent 0-dimensional tensors.
%
Scalars don't have as much structure as clearly only two structures apply.
%
Through these structures, scalars can also affect other tensors in crucial ways.
%
Constant propagation through arrays is known to be a complex compiler pass \cite{mcmichen_representing_nodate}.
%
Instead, we provide sparse scalars, which allow reductions but also specialize read accesses for the possible zero value.
%
We also provide early break scalars, which modify the stepper looplets to
re-specialize the loop whenever a reduction into an early break scalar hits an
annihilator.
%
We don't need to re-specialize other looplets because the stepper
is the only one which repeats a non-constant number of times.
%
We represent early break as a structural property rather than a program node
because it allows us to represent the tail of a loop where one scalar has hit an
annihilator but another scalar hasn't.
%
To our knowledge, sparse and early break scalars are novel contributions of this
work; other systems don't include them, limiting the impact of sparsity.

@@

The leaf level stores the actual entries of the tensor. In most cases, it is
sufficient to store each entry at a separate position in a vector. This is accomplished by the \textbf{ElementLevel}. However, when all of the
values are the same, an additional optimization can be made by storing the
identical value only once. In this work, we introduce the concept of a \textbf{PatternLevel}
 to handle this binary case. The PatternLevel has a fill value of $false$, and returning $true$ for all ``stored'' values. The PatternLevel allows us to easily represent unweighted graphs or other Boolean matrices.
